Comparing Reinforcement Learning and Finite State Machine Agents in Real Time Strategy Games: Impact on Player Experience

### 2025-04-20 Paper/main.tex
Almost done with milestone 2 and 3. forgot that LaTex does not use ** for bold like markdown files.

Literature review goes here

% Mention Csikszentmihalyi's flow theory and how it relates to player experience and immersion in games.

The goal of a game developer is to create a game that is fun and engaging, and they do this by creating a game that is challenging, but not too challenging.
As Csikszentmihalyi's flow theory states, the goal of a game is to create a state of flow, where the player is fully immersed in the game, and is not
bored, but also not overwhelmed. There are many ways to achieve this, and one of them is through the use of AI opponents. Traditionally, FSMs are
used as AI opponents, as they are simple to implement and easy to understand, and when done correctly, provide a good challenge to the player.
However, given enough time, players can learn the patterns of the FSMs, and exploit them, making the game feel boring. It is possible to combat this
through weakening the player, or making the AI more difficult, as is done in Souls-like games, however this can prove too challenging and overwhelm the player,
once again breaking the flow.

@@

\subsection{Research Questions}

1. How do RL and FSM AI opponents compare in terms of player experience in RTS games?
2. What are the key factors that influence player experience when playing against RL and FSM AI opponents in RTS games?
3. How do RL and FSM AI opponents impact player immersion in RTS games?

@@

% TODO:

@@

% TODO:

>>>>

The goal of a game developer is to create a game that is fun and engaging, and they achieve this by creating a game that is both challenging and rewarding, without being too difficult or frustrating.
This leads into the concept of player experience, which is a subjective measure of how much a player enjoys a game, and is influenced by many factors. One of the most influential frameworks for understanding player
experience is Csikszentmihalyi's flow theory \cite{csikszentmihalyi_flow_1990}, which describes a mental state in which a player is fully immersed, focused and involved in the game, leading to an improved sense of
enjoyment and intrinsic motivation. According to Csikszentmihalyi, flow occurs when the challenges presented match the player's skill level, and multiple conditions must be met for players to enter this state.
These are having clear goals, and immediate feedback, and when these conditions are met, players are more likely to lose track of time and become deeply immersed in the game world \cite{csikszentmihalyi_flow_1990}.

AI opponents play a critical role in maintaining this balance, as they provide the challenge and difficulty that the player must overcome. Traditionally FSMs are used for this, as they are simple to implement
and easy to understand, and when done correctly, provide a good challenge to the player. However, given enough time, players can learn the patterns of the FSMs, and exploit them,
making the game feel boring and leading to them falling out of the \textit{flow state}. It is possible to combat this through weakening the player, or making the AI more difficult,
as is done in Souls-like games, however this can prove too challenging and overwhelm the player, as well as making the percieved balance seem unfair, once again breaking the flow.

@@

\subsection{Research Questions}
The research questions for this study are:

\begin{enumerate}
    \item How do RL and FSM AI opponents compare in terms of player experience in RTS games?
    \item What are the key factors that influence player experience when playing against RL and FSM AI opponents in RTS games?
    \item How do RL and FSM AI opponents impact player immersion in RTS games?
\end{enumerate}

@@

This study adopts a \textbf{positivist research philosophy}, which means it emphasizes the use of objective measurements and observable phenomena, and it is suitable for this study as it aims
to evaluate the impact of RL and FSM AI opponents on player experience through measurable data, aligning it with \textbf{quantitative methodology}. It follows a \textbf{deductive approach}, with an \textbf{experimental research design},
as it starts with a hypothesis, and then using an experimental prototype, tests the hypothesis by allowing participants to play it, and then uses the data to observe the difference the independent variable (AI type)
has on the dependent variables \textbf(player experience, immersion, and perceived difficulty). This will then be used to answer the research questions and prove the hypothesis. It is important to note that the study
will also gather a very small amount of qualitative data, making it a \textbf{mixed-methods approach}. This is done to help align the more human aspects of player experience with the more quantitative aspects.

@@

\textbf{Validity}: The study ensures validity by desigining the experiment to control for confounding variables, such as the order in which participants play, since on the second playtest they would be more
familiar with the game mechanics and controls. The use of validating the survey results with in-game analytics and engagement metrics further streghtens the validity of the research being conducted.

\textbf{Reliability}: To help ensure reliability, the same setup, game environment and surveys will be used for all participants. Any and all instructions will be standardised, using text and/or video/audio
recordings to ensure that all participants are given the same instructions, without any bias that comes from the researcher. All datacollection processes will be automated, further standardising the process
and ensuring reliability.

\textbf{Generalizability/Transferability}: While the finding will be focused on the specific RTS game developed for this study, the insights gained can be generalised to other games within the RTS genre,
which follow similar game mechanics. The results may also be applicable to other game genres with similar AI opponent implementations, such as turn-based strategy games, but the transferability
of results could be limited, and as such should be explored in future research.
